Issues with Dynamical Measurements Across the Disk

### 2019-02-26 paper/ms.tex
revisions to 4.2

Suppose an open cluster of mass $m_c$ on an orbit with actions $\bm{J}_c$ is
disrupted. We would like to determine all stars in the {\em Gaia}
dataset\footnote{This is likely to be more successful with DR3/4 than DR2,
since upcoming data releases will contain a far greater number of stars with
radial velocities --- though targeted, follow-up radial velocity surveys may
allow this program to be performed sooner.} that are likely to be ejected
members from that cluster. One could generate an initial catalogue of
candidate members by integrating the cluster's orbit and selecting stars close
to that orbit, accounting for the fact that the stream will not exactly follow
the orbit of the cluster \citep[e.g.][]{2011MNRAS.413.1852E}.

One could then make a further selection by requiring that the action of each
field star $i$ be close to the cluster's actions to within some bound, i.e.
$\bm{J}_i - \bm{J}_c \in V_e$ where $V_e$ is some volume in action space
enclosing a region of acceptable error. A natural choice for $V_e$ is to
center it at the origin and allow its extent to be the intrinsic action-space
dispersion of a disrupted cluster. This dispersion can be estimated as
\citep[Section~8.3.3][]{2008gady.book.....B}
\beq \label{eq:action_disp}
\frac{\Delta J_i}{J_i} \sim \left(\frac{m_c}{M}\right)^{1/3}\text{,}
\eeq
where $m_c$ is the mass of the cluster and $M$ is the mass enclosed by the
cluster's orbit. A more sophisticated treatment of the action-space dispersion
of disrupted clusters can be obtained by following
\citet{2011MNRAS.413.1852E}, but such a treatment is premature for this work.

The act of comparing a field star's action to a cluster's action (i.e. in
computing $\bm{J}_i - \bm{J}_c$) is subject to significant error if the
cluster and field star have significantly different local midplanes and an
observer assumed a single global midplane.

To explore the magnitude of this effect, we consider the following program.
First, we assume a cluster mass $m_c$ and estimate the expected action-space
dispersion from Equation~\ref{eq:action_disp}. We then consult
Figure~\ref{fig:many_orbit_wrong_ref} to determine the $z$ offset necessary in
order for the induced error in the vertical action to be the same as the
expected action-space dispersion. If this were the case, an observer would
have to account for the different midplane of the field star and the cluster.
We perform this analysis for the thin-disk and thick-disk orbits considered in
Section~\ref{ssec:quant}, since the halo orbit is unrealistic for known open
clusters (Table~\ref{tab:real_clusters}).

With a characteristic $z$ offset in hand, we consult
Figure~\ref{fig:range_deltaphi} to convert this $z$ offset into a $\Delta
\phi$, which we convert to an arclength or characteristic distance, which we
refer to as the ``neighborhood radius'' or $R_{\n}$. Therefore, if an observer
is attempting to reconstruct a cluster with a given mass on a given orbit, we
will give a characteristic distance between a field star and the cluster
within which an observer might not need to consider differences in their
respective local midplanes.

@@

In Figure~\ref{fig:cluster_offset}, we show the $z$ offset necessary to induce
the action space dispersion computed from Equation~\ref{eq:action_disp}, using
the \texttt{gala} computed mass enclosed within a radius of $8.2\,\kpc$ ($\sim
10^{11}\,M_{\odot}$).

We now convert the $z$~offset-$m_c$ relation to an $R_{\n}$-$m_c$ relation,
where $R_{\n}$ indicates the neighborhood radius. An observer
attempting to reconstruct an open cluster of mass $m_c$ with perfect knowledge
of the midplane at the location of the cluster would expect to find the
ejected members using a dynamical cut for field stars within a distance
$R_{\n}$ of the cluster. If such an observer wants to find ejected stars that
are at a distance greater than $R_{\n}$, a more sophisticated analysis would
be required.

We use the relation in Figure~\ref{fig:range_deltaphi}, which shows the
typical $z$ offset at each $\Delta \phi$, to convert a given $z$~offset into a
typical $\Delta\phi$. We then convert this $\Delta \phi$ into a chord length
(see Figure~\ref{fig:fig_to_explain}) which we interpret as $R_{\n}$, the
neighborhood around the cluster. If the $z$~offset is greater than the maximum
range value of the galaxy, we simply report the maximum chord length
($2\times8.2\,\kpc$). The result of this procedure is shown in
Figure~\ref{fig:Rn_mc}. The solid lines indicate the neighborhood values
assuming the mean relation between $z$~offset and $m_c$, and the dashed lines
show the $+1\,\sigma$ and $-1\,\sigma$ relations, indicative of the strong
$\phi$-dependence of the midplane effect.

Figure~\ref{fig:Rn_mc} is most informative for the thin orbit (\thincolor).
The neighborhood around $\sim100\Msun$ clusters can be as small as
$\sim300\,\pc$ but up to $\sim1.2\,\kpc$ for \mi{} and \mf{}, though the
situation is marginally better for \mm{} with values ranging from
$\sim0.8\,\kpc$ to $\sim3\,\kpc$. The neighborhood values steadily increase
with $m_c$, topping out at $\sim5\,\kpc$ for $\sim10^4\,\Msun$ clusters in
\mi{} and \mf{}. In \mm{}, the neighborhood value reaches the maximum allowed of
$16.4\,\kpc$ for clusters of mass $\sim800\,\Msun$. The interpretation of this
is that the expected action space dispersion is so high for such high mass
clusters that the measured $z$~offset is not large enough throughout the
entire galaxy to be important.

For the thick-disk orbit (brown), the neighborhood is at the maximum value for
nearly all masses, except at the very low-mass end in \mi{}.

We repeat this entire procedure for each of the orbits of the real open
clusters in Table~\ref{tab:real_clusters} using the midplane variations from
\mi{}. We use \mi{} as a conservative choice, since it provides the greatest
midplane variation (Figure~\ref{fig:midplane}). We report the neighborhood
values for the mass of each cluster in the neighborhood column of
Table~\ref{tab:real_clusters}. Except for alphaPer, we find that the
neigbhorhood around each cluster is typically $\gtrsim 500\,\pc$. This
analysis assumes that the midplane from \mi{} can be accurately applied to
Milky Way orbits, which is not necessarily true. However, our analysis is
meant to be more illustrative than quantitative and these neighborhood numbers
should be understood as such.

>>>>

Suppose an open cluster of mass $m_c$ on an orbit with actions $\bm{J}_c$ is
disrupted. We would like to determine which stars with complete 6D phase-space positions, for example in the {\em Gaia}
dataset,\footnote{This is likely to be more successful with DR3/4 than DR2,
since upcoming data releases will contain a far greater number of stars with
radial velocities --- though targeted, follow-up radial velocity surveys may
allow this program to be performed sooner.} are likely to be ejected
members from that cluster. One could generate an initial catalogue of
candidate members by integrating the cluster's orbit and selecting stars close
to that orbit, accounting for the fact that the stream of ejected stars will not exactly follow
the orbit of the cluster \citep[e.g.][]{2011MNRAS.413.1852E}.

One could then make a further selection by requiring that the action of each
field star $i$ be close to the cluster's actions to within some bound, i.e.
$\bm{J}_i - \bm{J}_c \in V_e$, where $V_e$ is centered on $J_c$ and has size proportional to the intrinsic action-space dispersion of the disrupting cluster. This dispersion can be estimated as
\citep[Section~8.3.3 of][]{2008gady.book.....B}
\beq \label{eq:action_disp}
\frac{\Delta J_i}{J_i} \sim \left(\frac{m_c}{M}\right)^{1/3}\text{,}
\eeq
where $m_c$ is the mass of the cluster and $M$ is the mass enclosed by the
cluster's orbit around the Galaxy, in this case assumed to be the mass inside the Solar circle ($\sim 10^{11}\Msun$). Stars with actions within $V_e$ would then be considered cluster members.

%A more sophisticated treatment of the action-space dispersion
%of disrupted clusters can be obtained by following
%\citet{2011MNRAS.413.1852E}, but such a treatment is premature for this work.

Two issues are immediately apparent with this approach. The first is that it neglects the overall smooth distribution \emph{in action space} which will contribute some stars to the selection by chance. For dissolving open clusters in the thin disk, this background is not likely to be negligible, but we will assume for the sake of argument that it can be modeled within the ``dynamical solar neighborhood'' where it is safe to extend the local coordinate system. Since the effect of extending the coordinate frame beyond this region is generally to inflate the actions computed for thin-disk stars, using too large a volume is likely to under-estimate the density of the smooth thin-disk-like component in action space, while scattering more thin-disk stars onto thick-disk-like orbits and over-estimating the density of the smooth thick-disk-like component.

The second issue, which is related to the point of this paper, is that based on the results in our previous sections computing $\bm{J}_i$ for a field star and directly comparing it to  $\bm{J}_c$ will have significant systematic error if the cluster and field star are far enough apart to have significantly different local midplanes (see Figure \ref{fig:range_deltaphi}). Since our criterion for cluster membership (distance in action space) scales with the cluster mass, the region where the search will succeed varies with cluster mass as well. To illustrate this effect, we assume a cluster mass $m_c$ and estimate the expected action-space
dispersion from Equation~\ref{eq:action_disp}. We then consult
Figure~\ref{fig:many_orbit_wrong_ref} to determine the midplane $z$ offset necessary in
order for the induced error in the vertical action to be the same as the
expected action-space dispersion. We perform this analysis for the thin-disk-like and thick-disk-like orbits considered in
Section~\ref{ssec:quant}, assuming that the halo-like orbit is unlikely for members of known open
clusters (see Table~\ref{tab:real_clusters}). The result for the allowable midplane offset for these two orbits in shown in Figure \ref{fig:cluster_offset} as a function of cluster mass.

With a characteristic midplane offset in hand, we use the relation in Figure~\ref{fig:range_deltaphi}, which shows the typical midplane offset at each $\Delta \phi$, to convert a given midplane offset into a
typical $\Delta\phi$. As in this Figure we assume the cluster is at the Solar circle to obtain a physical scale, which we refer to as the ``neighborhood radius'' or $R_{\n}$. If the allowable midplane discrepancy is greater than the maximum excursions recorded in the simulated galaxy, we simply report the maximum chord length assuming an orbit on the Solar circle ($2\times8.2\,\kpc$). 
 If an observer is attempting to reconstruct a cluster with a given mass on a given orbit, $R_{\n}$ gives the characteristic search distance between a field star and the cluster within which an observer may not need to consider differences in their respective local midplanes when computing actions. 
The result is shown in
Figure~\ref{fig:Rn_mc}. The solid lines indicate the neighborhood values
assuming the mean relation between midplane offset and $m_c$, and the dashed lines
show the $+1\,\sigma$ and $-1\,\sigma$ relations based on the 50 fiducial Solar positions in each simulation (as discussed in Figure \ref{fig:range_deltaphi}). That this range is relatively wide illustrates the strong
$\phi$-dependence of the midplane offsets.

@@

%In Figure~\ref{fig:cluster_offset}, we show the $z$ offset necessary to induce
%the action space dispersion computed from Equation~\ref{eq:action_disp}, using
%the \texttt{gala} computed mass enclosed within a radius of $8.2\,\kpc$ ($\sim
%10^{11}\,M_{\odot}$).

%We now convert the $z$~offset-$m_c$ relation to an $R_{\n}$-$m_c$ relation,
%where $R_{\n}$ indicates the neighborhood radius. An observer
%attempting to reconstruct an open cluster of mass $m_c$ with perfect knowledge
%of the midplane at the location of the cluster would expect to find the
%ejected members using a dynamical cut for field stars within a distance
%$R_{\n}$ of the cluster. If such an observer wants to find ejected stars that
%are at a distance greater than $R_{\n}$, a more sophisticated analysis would
%be required.

Figure~\ref{fig:Rn_mc} shows that azimuthal midplane fluctuations will limit the volume over which actions can be used to match stars to open clusters on thin-disk-like orbits (\thincolor), depending on their mass. For thick-disk-like orbits (\thickcolor), the tolerances are much more forgiving: stars' actions can be compared over the entire Solar circle for nearly all cluster masses, except at the very low-mass end in \mi .
The neighborhood around $\sim100\Msun$ clusters can be as small as
$\sim300\,\pc$ but up to $\sim1.2\,\kpc$ for \mi{} and \mf{}, though the
situation is marginally better for \mm{} with values ranging from
$\sim0.8\,\kpc$ to $\sim3\,\kpc$. The neighborhood values steadily increase
with $m_c$, topping out at $\sim5\,\kpc$ for $\sim10^4\,\Msun$ clusters in
\mi{} and \mf{}. In \mm{}, the neighborhood size reaches the maximum allowed of
$16.4\,\kpc$ for clusters of mass $\sim800\,\Msun$. This figure also underlines that the degree to which such an analysis will succeed depends crucially on the expected degree of midplane fluctuations in the real Galaxy. If the Milky Way is more like \mi , the range of action-space analysis techniques like this is far more limited than if it more closely resembles \mm .

We also compute $R_{\n}$ for each of the orbits of the real open
clusters in Table~\ref{tab:real_clusters}, using the midplane variations from
\mi{} as a conservative choice, since it has the largest degree of
midplane variation (Figure~\ref{fig:midplane}). We report the neighborhood
values for the mass of each cluster in the neighborhood column of
Table~\ref{tab:real_clusters}. Except for alphaPer, we find that the
neigbhorhood around each cluster is typically $\gtrsim 500\,\pc$. This
analysis assumes that the MW resembles \mi{}, which is not necessarily true. However, this step of our analysis is meant to be more illustrative than quantitative: this calculation should actually be performed using real measurements of variations in the MW's stellar midplane. Such measurements should be possible in the near future with Gaia, with quality distance estimates the crucial limiting factor.
